---
title: "Adjusted correlations"
description: |
  A short description of the post.
author:
  - name: Shannon Pileggi
    url: {}
date: 09-09-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)   # general use ----
library(broom)       # tidying of stats results ----
library(performance) # to obtain marginal r-squared ----
library(lme4)        # linear mixed models ----   
```

# Data


```{r}
dat_raw <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv')
```

```{r}
dat_raw %>% count(playlist_genre)
```

```{r}
dat <- dat_raw %>% 
  mutate(year = str_sub(track_album_release_date, start = 1, end = 4) %>% as.numeric) %>% 
  filter(playlist_genre %in% c("pop", "rock", "rap")) %>%
  filter(year == 2019)
```


```{r}
dat %>% 
  dplyr::select_if(is.numeric) %>% 
  skimr::skim()
```

```{r}
dat %>% 
  dplyr::select_if(is.numeric) %>% 
  names(.)
```


# ggplot theme

Some personal ggplot preferences for use later.

```{r}
theme_shannon <- function(){
  # theme minimal creates white background and removes ticks on axes ----
  theme_minimal() +
  theme(
    # removes grid lines from plot ----
    panel.grid = element_blank(),
    # moves legend to top instead of side ----
    legend.position = "top",
    # removes title from legend, often superfluous ----
    legend.title = element_blank(),
    # creates the light gray box around the plot ----
    panel.background = element_rect(color = "#F2F2F2"),
    # creates the gray background boxes for faceting labels ----
    strip.background = element_rect(
      color = "#F2F2F2",
      fill = "#F2F2F2"
    ),
    # if using facet grid, this rotates the y text to be more readable ----
    strip.text.y = element_text(angle = 0),
    # this produces a fully left justified title ----
    plot.title.position = "plot"
  )
}
```




# Correlations

Correlations are a common technique used to assess the strength of the linear
association between two numeric variables. The main assumptions that go into this are:

1. The relationship is linear.

2. The observations are independent.

A common mistake that I see is when the request is to estimate the overall correlation, some 
may not to take into account important sources of variation in the data, which can negate
the independent observations assumption.


# Single feature

## Unadjusted correlations


For example, consider the three sports individually, yielding correlations between 
height and weight of 0.79, 0.46, and 0.73. One might expect an overall correlation 
between height and weight to be between 0.46 and 0.79; however, when estimated
in total ignoring sport, this yields an estimate of 0.81.

```{r}
dat %>% 
  bind_rows(dat %>% mutate(playlist_genre = "Total")) %>% 
  group_by(playlist_genre) %>% 
  nest() %>% 
  mutate(
    base_n = map_int(data, nrow),
    corr = map(data, ~ cor.test(x = .x$speechiness, y = .x$danceability) %>% broom::tidy())
    ) %>% 
  ungroup() %>% 
  unnest(c(data, corr)) %>% 
  mutate(
    corr_label =  glue::glue("{playlist_genre}\nn = {base_n}\n r = {scales::number(estimate, 0.01)}")
    #sport = fct_relevel(sport, "Shooting", "Gymnasitics", "Swimming", "Total")
    ) %>% 
  ggplot(aes(x = speechiness, y = danceability)) +
  geom_point(aes(color = corr_label), show.legend = FALSE) +
  facet_wrap(. ~ corr_label, ncol = 4) +
  #scale_color_manual(values = c("darkorange","purple","cyan4", "gray")) + 
  theme_shannon()
  
```

## Adjusted correlations

To estimate a correlation that adjusts for sport, implement a linear mixed model
with a fixed effect for height and a random effect for sport. The correlation
estimate adjusting for sport is the sign of the height coefficient multiplied by the
marginal r-squared of the model, which is the variance of the fixed effects.

```{r}
mixed_model <- lme4::lmer(danceability ~ speechiness + (1 | playlist_genre), dat)

height_sign <- mixed_model %>% 
  broom.mixed::tidy() %>% 
  filter(term == "speechiness") %>% 
  pull(estimate) %>% 
  sign()

# marginal rsq is the variance of fixed effects ----
adj_corr <- height_sign * sqrt(performance::r2(mixed_model)[["R2_marginal"]])

adj_corr
```

This yields a more sensible result for the overall correlation between height and weight.

# Many features

So far we estimated correlations between `danceability` and `speechiness` in three genres of music. 
Now let's examine the other numeric attributes.

```{r}
dat_stacked <- dat %>% 
  mutate(genre = playlist_genre, music_category = playlist_genre) %>% 
  bind_rows(dat %>% mutate(music_category = "total", genre = playlist_genre)) %>% 
  dplyr::select(track_id, music_category, genre, danceability, energy, 
                speechiness, acousticness, instrumentalness, liveness, valence) %>% 
  pivot_longer(
    -c(track_id, music_category, genre, danceability),
    names_to = "attribute",
    values_to = "value"
  ) 
```


```{r}
dat_stacked %>% 
  ggplot(aes(x = value, y = danceability)) +
  geom_point(aes(color = genre), show.legend = FALSE) +
  facet_grid(attribute ~ genre, scales = "free") +
  #scale_color_manual(values = c("darkorange","purple","cyan4", "gray")) + 
  theme_shannon()
```


```{r}
results_1 <- dat_stacked %>%
  group_by(music_category, attribute) %>% 
  nest() %>% 
  mutate(
    base_n = map_int(data, nrow),
    corr = map(data, ~ cor.test(x = .x$value, y = .x$danceability) %>% broom::tidy())
    ) %>% 
  ungroup() %>% 
  unnest(c(corr)) 
```

```{r}
compute_adj_corr <- function(data){
  
  mixed_model <- lme4::lmer(danceability ~ value + (1 | genre), data)

  height_sign <- mixed_model %>% 
    broom.mixed::tidy() %>% 
    filter(term == "value") %>% 
    pull(estimate) %>% 
    sign()

  # marginal rsq is the variance of fixed effects ----
  adj_corr <- height_sign * sqrt(performance::r2(mixed_model)[["R2_marginal"]])

  return(adj_corr)
}
```

```{r}
results_2 <- dat_stacked %>%
  filter(music_category == "total") %>% 
  group_by(attribute) %>% 
  nest() %>% 
  mutate(
    corr_adj = map_dbl(data, compute_adj_corr)
  )
```

