---
title: "Estimating correlations adjusted for group membership"
description: |
  A short description of the post.
author:
  - name: Shannon Pileggi
    url: {}
date: 09-09-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)      # general use ----
library(broom)          # tidying of stats results ----
library(performance)    # to obtain marginal r-squared ----
library(lme4)           # linear mixed models ----   
library(gt)             # create table ----
library(gtExtras)       # table formatting ----
library(palmerpenguins) # simpsons paradox example ----
```

# TL;DR

A common mistake that I see is when the request is to estimate the overall correlation, some 
may not to take into account important sources of variation in the data, which can negate
the independent observations assumption.

# Data

The data comes from the [TidyTuesday Spotify songs](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-01-21/readme.md){target="_blank"} data set.

```{r}
dat_raw <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv')
```


To reduce the scope and simplify, I am going to examine a reduced set of variables from `pop`, `rock`,
or `rap` songs in 2019.

```{r}
dat <- dat_raw %>% 
  mutate(year = str_sub(track_album_release_date, start = 1, end = 4) %>% as.numeric) %>% 
  filter(playlist_genre %in% c("pop", "rock", "rap")) %>%
  filter(year == 2019) %>% 
  dplyr::select(track_id, playlist_genre, danceability, energy, speechiness, acousticness, instrumentalness, liveness, valence) %>% 
  # create an additional variable for genre to be used later ----
  mutate(music_category = playlist_genre) 
```

```{r}
dat
```

Here are the definitions of the variables, straight from the TidyTuesday repo:

|variable                 |class     |description |
|:---|:---|:-----------|
|track_id                 |character | Song unique ID|
|playlist_genre           |character | Playlist genre |
|danceability             |double    | Danceability describes how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable and 1.0 is most danceable. |
|energy                   |double    | Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy. For example, death metal has high energy, while a Bach prelude scores low on the scale. Perceptual features contributing to this attribute include dynamic range, perceived loudness, timbre, onset rate, and general entropy. |
|speechiness              |double    | Speechiness detects the presence of spoken words in a track. The more exclusively speech-like the recording (e.g. talk show, audio book, poetry), the closer to 1.0 the attribute value. Values above 0.66 describe tracks that are probably made entirely of spoken words. Values between 0.33 and 0.66 describe tracks that may contain both music and speech, either in sections or layered, including such cases as rap music. Values below 0.33 most likely represent music and other non-speech-like tracks. |
|acousticness             |double    | A confidence measure from 0.0 to 1.0 of whether the track is acoustic. 1.0 represents high confidence the track is acoustic.|
|instrumentalness         |double    | Predicts whether a track contains no vocals. “Ooh” and “aah” sounds are treated as instrumental in this context. Rap or spoken word tracks are clearly “vocal”. The closer the instrumentalness value is to 1.0, the greater likelihood the track contains no vocal content. Values above 0.5 are intended to represent instrumental tracks, but confidence is higher as the value approaches 1.0. |
|liveness                 |double    | Detects the presence of an audience in the recording. Higher liveness values represent an increased probability that the track was performed live. A value above 0.8 provides strong likelihood that the track is live. |
|valence                  |double    | A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry). |

# ggplot theme

Some personal ggplot preferences for use later.

```{r}
theme_shannon <- function(){
  # theme minimal creates white background and removes ticks on axes ----
  theme_minimal() +
  theme(
    # removes grid lines from plot ----
    panel.grid = element_blank(),
    # moves legend to top instead of side ----
    legend.position = "top",
    # removes title from legend, often superfluous ----
    legend.title = element_blank(),
    # creates the light gray box around the plot ----
    panel.background = element_rect(color = "#F2F2F2"),
    # creates the gray background boxes for faceting labels ----
    strip.background = element_rect(
      color = "#F2F2F2",
      fill = "#F2F2F2"
    ),
    # if using facet grid, this rotates the y text to be more readable ----
    strip.text.y = element_text(angle = 0),
    # this produces a fully left justified title ----
    plot.title.position = "plot"
  )
}
```




# Correlations


```{r, echo = F}
set.seed(123)

#cmat <- c(   1,   0.99, 0, -0.99,
#          0.99,   1,    0,     0,
#             0,   0,    1,     0,
#         -0.99,   0,    0,     1)
#
#dat0 <- faux::rnorm_multi(n = 100, 
#                  mu = 0.5,
#                  r = cmat, 
#                  varnames = c("y", "x1", "x2", "x3"),
#                  empirical = FALSE)

dat1 <- faux::rnorm_multi(n = 100, 
                  mu = c(0.5, 0.5),
                  sd = c(0.15, 0.15),
                  r = 0.99, 
                  varnames = c("y", "x"),
                  empirical = TRUE)

dat2 <- faux::rnorm_multi(n = 100, 
                  mu = c(0.5, 0.5),
                  sd = c(0.15, 0.15),
                  r = 0, 
                  varnames = c("y", "x"),
                  empirical = TRUE)

dat3 <- faux::rnorm_multi(n = 100, 
                  mu = c(0.5, 0.5),
                  sd = c(0.15, 0.15),
                  r = -0.99, 
                  varnames = c("y", "x"),
                  empirical = TRUE)
```

```{r, echo = FALSE}
#devtools::install_github("shannonpileggi/ralph")
dat_all <- dat1 %>% mutate(feature = 1) %>% 
  bind_rows(dat2 %>% mutate(feature = 2)) %>% 
  bind_rows(dat3 %>% mutate(feature = 3)) %>% 
  rename(danceability = y, value = x) %>% 
  group_by(feature) %>% 
  nest() %>% 
  mutate(results = purrr::map(.x = data,.f = ralph::compute_corr, var1 = danceability, var2 = value)) %>% 
unnest(c(data, results)) %>% 
  ungroup() %>% 
  mutate(label = glue::glue("Feature {feature}\n r = {scales::number(correlation)}"))
```

Correlations (_r_) take on values between -1 and 1, and measure the strength of the linear association between two variables. Here are the three extreme forms of correlation:

```{r, echo=FALSE}
dat_all %>% 
  ggplot(aes(x = danceability, y = value)) +
  geom_point(color = "#E76F51", alpha = 0.7) +
  facet_wrap(. ~ label) +
  theme_shannon()
```

The two main assumptions that go into correlations are:

1. The relationship is linear.

2. The observations are independent.

The linearity of the relationship can be evaluated by examining a scatter plot. Independence of observations is evaluated by thinking about the origin and nature of the data. A classic way of violating the independence assumption is when observations arise from repeated measures; a sneakier way the independence observation can be violated is from what Isabella  Ghement refers to as a [random grouping factor](https://twitter.com/IsabellaGhement/status/1443773295426359298){target="_blank"}. 

Let's check this out by examining correlations on a single feature in the spotify data set. `r emo::ji("monocle")`


# Single feature

## Unadjusted correlations


For example, consider the three sports individually, yielding correlations between 
height and weight of 0.79, 0.46, and 0.73. One might expect an overall correlation 
between height and weight to be between 0.46 and 0.79; however, when estimated
in total ignoring sport, this yields an estimate of 0.81.

```{r}
# genre colors
colors_genre <- c(
  "pop" = "#E76F51", 
  "rock" = "#264653", 
  "rap" = "#E9C46A"
)
```



```{r}
dat %>% 
  bind_rows(dat %>% mutate(music_category = "total")) %>% 
  group_by(music_category) %>% 
  nest() %>% 
  mutate(
    base_n = map_int(data, nrow),
    corr = map(data, ~ cor.test(x = .x$speechiness, y = .x$danceability) %>% broom::tidy())
    ) %>% 
  ungroup() %>% 
  unnest(c(data, corr)) %>% 
  mutate(
    corr_label =  glue::glue("{music_category}\nn = {base_n}\n r = {scales::number(estimate, 0.01)}")
    #sport = fct_relevel(sport, "Shooting", "Gymnasitics", "Swimming", "Total")
    ) %>% 
  ggplot(aes(x = speechiness, y = danceability)) +
  geom_point(aes(color = playlist_genre), alpha = 0.2, show.legend = FALSE) +
  facet_wrap(. ~ corr_label, ncol = 4) +
  scale_color_manual(values = colors_genre) + 
  theme_shannon()
  
```

Here, we can see that the correlations between `speechiness` and `danceability` are 
0.06 (`pop`), 0.13 (`rap`), and -0.11 (`rock`), and so it might be expected that
the overall correlation between `speechiness` and `danceability` would be between `-0.11` 
and `0.13`, yet it is estimated to be `0.25`, which is stronger than any individual
correlation. `r emo::ji("thinking")`

## Adjusted correlations

To estimate a correlation that adjusts for genre, we can implement a linear mixed model
with a fixed effect for `speechiness` and a random effect for `genre`. The correlation
estimate adjusting for `genre` is the sign of the `speechiness` coefficient multiplied by the
square root of the marginal r-squared of the model, which is the variance of the fixed effects.

```{r}
mixed_model <- lme4::lmer(danceability ~ speechiness + (1 | playlist_genre), dat)

height_sign <- mixed_model %>% 
  broom.mixed::tidy() %>% 
  filter(term == "speechiness") %>% 
  pull(estimate) %>% 
  sign()

# marginal rsq is the variance of fixed effects ----
adj_corr <- height_sign * sqrt(performance::r2(mixed_model)[["R2_marginal"]])

adj_corr
```

This yields a more sensible result for the overall correlation between height and weight, and
passes the sniff test that the correlation should reasonably be between `-0.11` 
and `0.13`.

# Many features

So far we estimated correlations between `danceability` and `speechiness` in three genres of music. 
Now let's examine the other numeric attributes.

```{r}
dat_stacked <- dat %>% 
  bind_rows(dat %>% mutate(music_category = "total")) %>% 
  dplyr::select(track_id, music_category, playlist_genre, danceability, energy, 
                speechiness, acousticness, instrumentalness, liveness, valence) %>% 
  pivot_longer(
    -c(track_id, music_category, playlist_genre, danceability),
    names_to = "feature",
    values_to = "value"
  ) #%>% 
  # reordered by a sneak peak at results below ----
  # mutate(feature = fct_relevel(feature, "valence", "speechiness", "energy", "acousticness", "liveness", "energy"))
```


```{r}
dat_stacked %>% 
  ggplot(aes(x = value, y = danceability)) +
  geom_point(aes(color = playlist_genre), alpha = 0.2, show.legend = FALSE) +
  facet_grid(feature ~ music_category, scales = "free") +
  scale_color_manual(values = colors_genre) +
  theme_shannon()
```


```{r}
results_1 <- dat_stacked %>%
  group_by(music_category, feature) %>% 
  nest() %>% 
  mutate(
    corr = map(data, ~ cor.test(x = .x$value, y = .x$danceability) %>% broom::tidy())
    ) %>% 
  ungroup() %>% 
  unnest(c(corr)) %>% 
  dplyr::select(music_category, feature, corr = estimate)
```

```{r}
compute_adj_corr <- function(data, var_dep, var_ind, var_group){
  
  #mixed_formula <- "danceability ~ value + (1 | playlist_genre)"
  mixed_formula <- glue::glue("{var_dep} ~ {var_ind} + (1 | {var_group})")
  
  mixed_model <- lme4::lmer(mixed_formula, data)

  height_sign <- mixed_model %>% 
    broom.mixed::tidy() %>% 
    filter(term == var_ind) %>% 
    pull(estimate) %>% 
    sign()

  # marginal rsq is the variance of fixed effects ----
  adj_corr <- height_sign * sqrt(performance::r2(mixed_model)[["R2_marginal"]])

  return(adj_corr)
}
```


```{r}
results_2 <- dat_stacked %>%
  filter(music_category == "total") %>% 
  group_by(feature) %>% 
  nest() %>% 
  mutate(
    corr = map_dbl(data, compute_adj_corr, var_dep = "danceability", var_ind = "value", var_group = "playlist_genre")
  ) %>% 
  mutate(music_category = "total adj") %>% 
  dplyr::select(music_category, feature, corr)
```

```{r}
results_all <- results_1 %>% 
  bind_rows(results_2) %>% 
  pivot_wider(
    names_from = "music_category",
    values_from = "corr"
    ) %>% 
  mutate_if(is.numeric, round, 2) %>% 
  arrange(desc(`total adj`))
```

```{r}
results_all %>% 
  gt::gt() %>% 
  gt_hulk_col_numeric(pop:`total adj`)
```


# Simpson's paradox

```{r}
dat_penguins <- penguins %>% 
  dplyr::select(species, bill_depth_mm, body_mass_g) %>% 
  # duplicate species variable for coloring & grouping ---
  mutate(species_category = species) %>% 
  drop_na()
```

```{r}
colors_penguins <- c(
  "Adelie" = "darkorange",
  "Chinstrap" = "purple",
  "Gentoo" = "cyan4"
  )
```


```{r}
dat_penguins %>% 
  bind_rows(dat_penguins %>% mutate(species_category = "All")) %>% 
  group_by(species_category) %>% 
  nest() %>% 
  mutate(
    base_n = map_int(data, nrow),
    corr = map(data, ~ cor.test(x = .x$bill_depth_mm, y = .x$body_mass_g) %>% broom::tidy())
    ) %>% 
  ungroup() %>% 
  unnest(c(data, corr)) %>% 
  mutate(
    species_category = fct_relevel(species_category, "Gentoo", "Chinstrap", "Adelie", "All"),
    corr_label =  glue::glue("{species_category}\nn = {base_n}\n r = {scales::number(estimate, 0.01)}"),
    corr_label = fct_reorder(as.character(corr_label), as.numeric(species_category))
   ) %>% 
  ggplot(aes(x = bill_depth_mm, y = body_mass_g)) +
  geom_point(aes(color = species), alpha = 0.2, show.legend = FALSE) +
  geom_smooth(method = "lm", color = "darkgray", se = FALSE) +
  facet_wrap(. ~ corr_label, ncol = 4) +
  scale_color_manual(values = colors_penguins) + 
  theme_shannon()
```

```{r}
compute_adj_corr(
  data = dat_penguins, 
  var_dep = "body_mass_g", 
  var_ind = "bill_depth_mm", 
  var_group = "species"
  )
```

