---
output: github_document
---


<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message=FALSE, 
  warning=FALSE
)
```

# About

A personal website:

⚒️ created with the [distill](https://rstudio.github.io/distill/) package in R,

💡 continuously inspired by community contributed tips at the [distillery](https://jhelvy.github.io/distillery/) and all content creators out there,

🗨️ with commenting by [giscus](https://giscus.app/), and

🚀 deployed through Netlify.


```{r}
library(magrittr)
library(stringr)
library(dplyr)
library(xml2)
library(lubridate)
library(ggplot2)
```

```{r}
# inspired by Matt Dray's blog 
# https://www.rostrum.blog/2021/04/14/gha-readme/
xml <- read_xml("https://www.pipinghotdata.com/blog.xml")

titles <- xml_find_all(xml, ".//title") %>% 
  str_remove_all("<title>|</title>") %>% 
  .[-1]

links <- xml_find_all(xml, ".//link") %>% 
    str_remove_all("<link>|</link>") %>% 
  .[-1]

dates <- links %>% 
  str_remove(".*posts/") %>% 
  str_sub(1, 10)

                      # extract text

# data frame of all posts
posts <- tibble(dates, titles, links)

# number of days since last post
days_last_post <- Sys.Date() - max(lubridate::as_date(dates))

posts_since_2020 <- posts %>% 
  mutate(dates = lubridate::as_date(dates)) %>% 
  filter(dates >= lubridate::as_date("2020-01-01"))

posts_per_month <- scales::number((nrow(posts_since_2020) / (as.numeric(today() - min(posts_since_2020$dates)))) * 30, 0.1)

days_bt_posts <- scales::number((as.numeric(today() - min(posts_since_2020$dates))) / nrow(posts), 1)

```




# Some stats ⚠️ WIP

🎉 Piping Hot Data has **`r nrow(posts)`** posts since **`r min(lubridate::as_date(dates))`**!

📅 That's a post roughly every **`r days_bt_posts`** days, or about `r posts_per_month` posts per month, since `r min(posts_since_2020$dates)`.


✍️ The last post was published **`r days_last_post`** days ago ([`r titles[1]`](`r links[1]`)). Your feedback motivates me to keep writing. 😉

```{r plot-obj, fig.height=1, fig.width=10}
# Create plot object
posts %>%
  ggplot(aes(x = lubridate::as_date(dates), y = 1)) +
  geom_point(shape = "|", size = 5,   color = "#f95355") + 
  theme(
    panel.grid = element_blank(),
    # creates the light gray box around the plot ----
    panel.background = element_rect(fill = "white", color = "white"),
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_text(size = 14),
    axis.ticks = element_blank(),
    plot.margin = margin(0, 10, 0, 10)
  ) +
  ggtitle("Published posts")
```



<details><summary>`r emo::ji("open_file_folder")` Click to expand a full list of posts</summary>

```{r posts-table}
posts %>%
  transmute(
    Date = dates, 
    Title = paste0("[", titles, "](", links, ")")
  ) %>% 
  knitr::kable()
```
</details>

---

_The automatically rendered readme was inspired by [Matt Dray](https://www.rostrum.blog/2021/04/14/gha-readme/); giscus commenting
was motivated by [Joel Nitta](https://www.joelnitta.com/posts/2021-11-24_using-giscus/)._


_Updated `r Sys.time()`_


```{r, eval=FALSE}
# https://tgerke.github.io/github-actions-with-r/#1
# https://github.com/r-lib/actions/issues/186
# https://github.com/r-lib/actions/tree/master/setup-r-dependencies
```

