---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message=FALSE, 
  warning=FALSE
)
```

# About

A personal website 

⚒️ created with the [distill](https://rstudio.github.io/distill/) package in R

💡 continuously inspired by community contributed tips at the [distillery](https://jhelvy.github.io/distillery/) and all content creators out there

🗨️ with commenting by [giscus](https://giscus.app/)

🚀 deployed through Netlify


```{r}
library(tidyverse)
library(xml2)
library(lubridate)
```

```{r}
# inspired by Matt Dray's blog 
# https://www.rostrum.blog/2021/04/14/gha-readme/
xml <- read_xml("https://www.pipinghotdata.com/blog.xml")

titles <- xml_find_all(xml, ".//title") %>% 
  str_remove_all("<title>|</title>") %>% 
  .[-1]

links <- xml_find_all(xml, ".//link") %>% 
    str_remove_all("<link>|</link>") %>% 
  .[-1]

dates <- links %>% 
  str_remove(".*posts/") %>% 
  str_sub(1, 10)

                      # extract text

# data frame of all posts
posts <- tibble(dates, titles, links)

# number of days since last post
days_last_post <- Sys.Date() - max(lubridate::as_date(dates))

```




# Some stats

🎉 Piping Hot Data has **`r nrow(posts)`** posts since **`r min(lubridate::as_date(dates))`**!



✍️ The last post was published `r days_last_post` days ago ([`r titles[1]`](`r links[1]`)). Your feedback motivates me to keep writing. 😉

```{r plot-obj, fig.height=1, fig.width=10}
# Create plot object
posts %>%
  ggplot(aes(x = lubridate::as_date(dates), y = 1)) +
  geom_point(shape = "|", size = 5,   color = "#f95355") + 
  theme(
    panel.grid = element_blank(),
    # creates the light gray box around the plot ----
    panel.background = element_rect(fill = "white", color = "white"),
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_text(size = 14),
    axis.ticks = element_blank(),
    plot.margin = margin(0, 10, 0, 10)
  ) +
  ggtitle("Published posts")
```



<details><summary>`r emo::ji("open_file_folder")` Click to expand a full list of posts</summary>

```{r posts-table}
posts %>%
  transmute(
    Date = dates, 
    Title = paste0("[", titles, "](", links, ")")
  ) %>% 
  knitr::kable()
```
</details>

_This automatically rendered readme was inspired by [Matt Dray](https://www.rostrum.blog/2021/04/14/gha-readme/)._

---

_Updated `r Sys.time()`_
