---
title: "Your first R package in 1 hour"
description: |
  Tools that make R package development easy
base_url: https://www.pipinghotdata.com/
preview: workshop_thumnnail3.jpg
twitter:
  site: "@PipingHotData"
  creator: "@PipingHotData"
date: 11-05-2020
author:
  - name: Shannon Pileggi
output:
  distill::distill_article:
    toc: true
    toc_depth: 1
    self_contained: false
draft: true
creative_commons: CC BY
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# TL; DR

This blog post accompanies the [R-Ladies Philly](https://www.rladiesphilly.org/) workshop on November 12, 2020.  
Central idea: leverage [devtools](https://github.com/r-lib/devtools) and [usethis](https://usethis.r-lib.org/) in package development!

# Abstract

This workshop is for people looking to learn how to make their own R packages and learn how to use “usethis” and “devtools” for package development. The workshop will cover handy one time functions (i.e., `use_this::create_package`) as well as functions used continuously throughout package development (i.e., `devtools::document`). At the end of the hour you should have a working, well-documented package with a single function, as well as a better understanding of the files and file structure required for R packages.

This workshop is suitable for beginner to intermediate R users. Attendees should be familiar with functions, but will not be writing their own function in this workshop. Familiarity with pipe or tidyverse is helpful.

# Set up

This material was developed using:

| Software / package  | Version               |
|---------------------|-----------------------|
| R                   | 4.0.2                 | 
| RStudio             | 1.3.1073              | 
| `devtools`          | 2.3.2                 | 
| `usethis`           | 1.6.3                 |
| `tidyverse`         | 1.3.0                 |
| `broom`             | 

Please update all software / packages prior to following along with this tutorial, as otherwise errors may arise.

# Function reference

::: l-body-outset
| Usage    | Function   | Purpose               |
|----------|------------|-----------------------|
| Single   | `usethis::create_package("path/packagename")`  | initialize package   | 
|          | `usethis::use_mit_license("Your name")`        | adds license         |
|          | `usethis::use_pipe()`                          | add pipe function as a dependency |
| Multiple | `devtools::check()`           | build and check package |
|          | `usethis::use_r("file-name")` | create R script for function |
|          | `usethis::use_package("package-name")` | add package dependency |
|          | `devtools::document()` | document a function |
:::

# Copy and paste materials

## Function

The `compute_corr` function is a wrapper for `cor.test` to produce tidy output for Pearson's correlation estimate (along with a p-value) for the relationship between two quantitative variables.

```{r include=FALSE}
library(tidyverse)
```


```{r define-function, eval = TRUE}
compute_corr <- function(data, xvar, yvar){
  
  cor.test(x = data[[xvar]], y = data[[yvar]]) %>% 
    # tidy up results ----
    broom::tidy() %>% 
    # retain and rename relevant bits ----
    dplyr::select(
      correlation = estimate, 
      pval = p.value
    )
  
}
```

Here is an example execution:

```{r}
compute_corr(mpg, "cty", "hwy")
```
## Documentation


# First package

Open an R session and submit the following, modified to your desired location.  Here,
I am creating a package named `ralph` on my desktop. (This name is inspired by **R**-**L**adies **Ph**illy.) 

```{r create-package, eval=FALSE}
usethis::create_package("C:/Users/Shannon.Pileggi/Desktop/ralph")
```

Now you have a package!  


# First check

# First function


# Summary of steps
