---
title: "curating"
description: |
  A short description of the post.
base_url: https://www.pipinghotdata.com/
preview: 
twitter:
  site: "@PipingHotData"
  creator: "@PipingHotData"
date: 03-14-2021
author:
  - name: Shannon Pileggi
output:
  distill::distill_article:
    toc: true
    toc_depth: 1
    self_contained: false
draft: true
creative_commons: CC BY
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
folder <- "2021-03-14-curating"
```

```{r}
library(tidyverse)
library(ggtext)
library(ggiraph)
```


working title:

working description:

preview image: maybe a cute behind the curtain?

# TL; DR

# Signing up

R-Ladies has a [comprehensive guide](https://guide.rladies.org/rocur/about/){target="_blank"} describing why the program exists, procedures and protocols for the week, and 
tips for successful curation. You sign up by [submitting a form](https://docs.google.com/forms/d/e/1FAIpQLSevtboT8TNjz87qe5HZzuBHuY0aNteaZNF5e2xFQXtSeU0c0w/viewform){target="_blank"} - give yourself at least 30 minutes to sign up as part of the
form includes filling out details that complete your [curating profile](https://twitter.com/WeAreRLadies/status/1361139211819180032){target="_blank"}.

# Date selection

You can view the [schedule](https://docs.google.com/spreadsheets/d/1pmphnR9EinGuAKbVqGzMIt6cWYmHNbeFVzaKJB59JdY/edit#gid=1429703967){target="_blank"} of previous and upcoming curators to identify available dates.  

Being a curator will be time intensive, so be kind to yourself. Choose dates when you will have time to invest and a flexible work schedule. I chose Feb 15-20 
because I hoped by then I should be recovered from a very intense Q4 work cycle;
additionally, Feb 15 (President's Day) was a company holiday.

An additional consideration is to schedule your curation to coincide with dates that align with your interests. For example, are you passionate about Black History Month in February, LGBT Pride Month in June, or Universal Human Rights Month in December? 
If so, take advantage of the `@WeAreRLadies` large platform as an opportunity to inform and educate others on issues that are important to you as they relate to the 
R community.

# Timeline

_This section needs more work._

_1. Graphical or table presentation?_

_2. For table - format with inline options vs knitr-esqe options?_

_3. Also not shown - noted brief content ideas starting when I signed up to the day of curation._


```{r}
timeline <- read_csv(here::here("_posts", folder, "data", "rladies_timeline.csv")) %>% 
  mutate(
    even_odd = (id %% 2) == 0,
    label = glue::glue("*{time_frame} before*<br>{str_wrap(action, 10)}")
    )
```

Here is what I did leading up to volunteering for the curator role. 

```{r}
timeline %>% 
  dplyr::select(time_frame, action)
```

I became more active on twitter in 2018, but I didn't sign up to curate until 2021. 
For me, I needed to learn how to interact on twitter and what was well-received before I had the confidence to share on a larger platform. 

When I signed up to curate, I selected a date three months out to give myself time to prepare content. 

Speaking with my manager was both important and rewarding. First, I wanted to cover
my bases in case corporate big brother noticed an unusual spike in my twitter activity
during working hours. Moreover, my manager had no idea that R-Ladies even existed, and she was enthusiastic and supportive. Lastly, I did discuss potential content with 
her that touched on work topics. One idea was given the green light and another 
was reasonably denied. I am very glad we had this honest conversation.

This is not necessarily a recommended or fixed timeline - this is just what worked 
for me. 

_Alternative presentation: I don't think this figure has much promise, maybe
just stick to table format._

```{r}
timeline %>% 
  ggplot() +
  geom_segment(aes(x = 1, xend = 7, y = 0, yend = 0), size = 0.5) +
  geom_segment(aes(x = id, xend = id, y = 0, yend = ifelse(even_odd, -.3, 0.3)), size = 0.3) +
   geom_richtext(
    aes(
      x     = id,
      y     = ifelse(even_odd, -.3, 0.3),
      label = label,
      #vjust = "top",
      #hjust = ifelse(long_next < -100, 0, 1),
    ))
```



# Content inspiration

As I approached my curating week, I recalled previous `@WeAreRLadies` that were
memorable for me, my previous experience as an educator, and some reflection
questions to inspire content.

| Inspiration  source | Example realization   |
|---------------------|-----------------------|
| 1. Rotating curator Mine Çetinkaya-Rundel ([`@minebocek`](https://twitter.com/minebocek){target="_blank"}) tweets awesome [gifs](https://twitter.com/WeAreRLadies/status/1064704918102163457){target="_blank"}| Gifs for R code demos like [this one](https://twitter.com/WeAreRLadies/status/1361802517735178243){target="_blank"}  |
| 2. Rotating curator Megan Stodel ([`@MeganStodel`](https://twitter.com/MeganStodel){target="_blank"}) tweets a [project](https://twitter.com/WeAreRLadies/status/1313177623128944645){target="_blank"} inspired by curating | An R project to [introduce myself](https://twitter.com/WeAreRLadies/status/1361286341317779456){target="_blank"} as a curator |
| 3. Rotating curator Julia Piaskowski ([`@SeedsAndBreeds`](https://twitter.com/SeedsAndBreeds){target="_blank"}) tweets a great technical thread on [ANOVA](https://twitter.com/WeAreRLadies/status/1223790298024726528){target="_blank"} | A thread on [blogging resources](https://twitter.com/WeAreRLadies/status/1363144545677017089){target="_blank"} |
| 4. Prior experience as an educator | Starting discussion with a [question](https://twitter.com/WeAreRLadies/status/1361332603274612739){target="_blank"} |
| 5. What am I passionate about lately? | [Blogging](https://twitter.com/WeAreRLadies/status/1363144545677017089){target="_blank"} |
| 6. What did I have to overcome to be where I am today? | Learning how to [ask questions online](https://twitter.com/WeAreRLadies/status/1362370580708790274){target="_blank"} |
| 7. What have colleagues or students asked me about? | What needs [updating](https://twitter.com/WeAreRLadies/status/1362114431090573315){target="_blank"}  and when |
| 8. What are some R functions or packages that have helped me recently? | [sortable ](https://twitter.com/WeAreRLadies/status/1361418527870132226){target="_blank"} package |
| 9. What are R-Ladies voices I can amplify while I have this large platform? | Re-tweeting recognition like [this](https://twitter.com/lisalendway/status/1363159170531401729){target="_blank"} |

# Drafting content

Leading up to my curation week, I regularly jotted down brief notes of content ideas.
The week before curation, I started fleshing out those ideas into actual tweets and actually wrote them down in a document. Not all of my ideas ended up in a draft,
and rarely did the draft get tweeted out exactly as I had written. 

One challenge with drafting tweets in a document was being mindful of character limits
and anticipating where the breaks would be for threads. I actually started copying into a send tweet window to preview and then pasting it back into my draft document. There is software that facilitates drafting tweets - for example, someone recommended
 [getchirrapp.com](https://getchirrapp.com/){target="_blank"} to me the week of my curation, but I didn't end up using it.

# Polls

As I already mentioned, I love starting discussion with questions! I planned three polls and follow-up discussion to happen throughtout the week. Polls on twitter 
are open for 24 hours and only allow up to four response options. 

```{r}
polls <- read_csv(here::here("_posts", folder, "data", "rladies_polls.csv")) %>% 
  mutate(
    day = factor(day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday",
                             "Friday", "Saturday")),
    poll_label = glue::glue("Poll #{poll}") %>% fct_reorder(poll),
    tweet_label = glue::glue("Tweet #{step}") %>% fct_reorder(step) %>% fct_rev(),
    step = factor(step) %>% fct_rev() %>% fct_drop(),
    content = str_wrap(content, 12),
    content_label = glue::glue("{type}:\n{content}"),
    content_click = sprintf("window.open(\"%s%s\")", url, ""),
    content_url = 
       str_c("<a href='", content, "' target='_blank'>", url, "</a>")
    )
```


```{r}
#tooltip_css <- "background-color:gray;color:white;padding:10px;border-radius:5px;"
```


```{r}
p <- polls %>% 
  ggplot(aes(x = 1, y = tweet_label, fill = poll_label, tooltip = embed, onclick = content_click)) +
  facet_wrap(. ~ day, ncol = 6, drop = F) +
  geom_tile_interactive(color = "white") +
  geom_text(aes(label = content_label), size = 2.5) +
  theme_minimal() +
  theme(
    # removes grid lines from plot ----
    panel.grid = element_blank(),
    # moves legend to top instead of side ----
    legend.position = "top",
    # removes title from legend, often superfluous ----
    legend.title = element_blank(),
    # creates the light gray box around the plot, may or may not want this ----
    #panel.background = element_rect(color = "#F2F2F2"),
    # creates the gray background boxes for faceting labels ----
    strip.background = element_rect(
      color = "#F2F2F2",
      fill = "#F2F2F2"
    ),
    # this produces a fully left justified title ----
    plot.title.position = "plot",
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    #axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    )
```

_What do you think of this graphical presentation? The graphic has both hover enabled to show the tweek and click enabled to take you to the tweet. The hover over was taken from embed tweet on twitter and runs off the page - I would need to make adjustments
here for readability if I keep.  Also need a custom color scheme._


```{r}
girafe(ggobj = p,
       #opts_tooltip(use_fill = TRUE)
       #opts_tooltip(css = tooltip_css, use_fill = TRUE)
       )
```





# First and last tweets

The introduction and farewell tweets as a curator are important as this is when you actually tell people your name or personal twitter handle. I have seen 
many excellent curators fall short on engagement (likes and re-tweets) with their first and last tweets, resulting in less than merited recognition for the wonderful curator. 

To address this, I aimed to create content-rich [first](https://twitter.com/WeAreRLadies/status/1361286341317779456){target="_blank"} and [last](https://twitter.com/WeAreRLadies/status/1363286037846511616){target="_blank"} tweets to give twitter users more motivation to like or re-tweet, and I also connected the content with links to my blog so that users could easily learn more about me.

# TweetDeck

When you serve as a curator, you will be tweeting from [TweetDeck](https://tweetdeck.twitter.com/){target="_blank"}, and it is hard to separate curator experience from the technology. Tweeting from TweetDeck is a different twitter interface and experience, and feels like

_find right gif - looking for someone standing in front of screen or window with code all around them_

TweetDeck had a learning curve for me, and there were limitations to the platform
that added challenges to curating, which included:

1. There was no `+` enabled to easily create threads (I had to send tweet and then comment on tweet for a thread; and it was initially hard to make sure thread appeared in correct order).

2. Consequently, I could not draft and save threads in TweetDeck to send later.

3. I could not send polls from the curator account on TweetDeck; I had to send polls from my personal account and then re-tweet from the curator account.

4. Depending on the content already in the send tweet interface, sometimes other options in TweetDeck would disappear, like the emoji, gifs, and upload image buttons. I kept [emojipedia](https://emojipedia.org/){target="_blank"} open to easily copy and paste emojis into my tweets, and it took trial and error to get everything I wanted in a single tweet.

5. When uploading local content, you can add [descriptions](https://help.twitter.com/en/using-twitter/picture-descriptions){target="_blank"} to both gifs and images in the regular twitter interface to create inclusive content for community members that use assistive reading technology; however, in TweetDeck, descriptions were enabled for images and not gifs.

I spent a lot of time my first couple of days as a curator getting used to TweetDeck and researching alternative solutions and plug-ins that ultimately did not help. Twitter is targeting TweetDeck [enhancements](https://www.theverge.com/2021/3/9/22321991/twitter-tweetdeck-overhaul-redesign-product-changes){target="_blank"} later in 2021, so I don't think it is worth documenting all of my methods and work-arounds. However, if you are serving as a curator and struggling with TweetDeck, please reach out - I am happy to share what ended up working for me.

# Reflection

_this is getting long, break into multiple sections?_

Just because a twitter account has >20K followers, the likes, re-tweets, and comments don't come automatically. You still have to earn engagement with your content. Many of the tweets I sent had little engagement, and that is OK. Supporting your curators by engaging with their tweets or sending notes of encouragement is _much_ appreciated. I thank everyone who engaged with me during my curation, with a special shout out to Alison Hill who re-energized me mid-week with comments on the R-Ladies bloggers [thread](https://twitter.com/WeAreRLadies/status/1362021673239785473){target="_blank"}.

My tweets were not perfect, and that is also OK. I messed up threads, had typos, and shared deprecated code, among other things. Check out my [blooper reel](https://twitter.com/PipingHotData/status/1364183660744896513){target="_blank"} for tweets that I bungled.

One regret that I do have from my curating week is failing to explicitly knowledge Black History Month as I was tweeting in February. I wish had prepared at least one tweet or better amplified the voices of black members of the R community while I had the large platform. 

Serving as a curator was intimidating and time consuming, but I am very glad I did it. Many good things have happened as direct result of that week, including: 

* [discussing](https://twitter.com/WeAreRLadies/status/1363144545677017089){target="_blank"} comparisons between blogdown and distill with Alison Hill 

* [collaborating](https://www.pipinghotdata.com/posts/2021-03-08-r-ladies-styled-code-gifs-with-xaringan-and-flipbookr/){target="_blank"} with Silvia Canelón to style code gifs

* engaging with new people on twitter that I want to continue to engage with

* learning about valuable new-to-me packages, functions, and work flows 

* being invited to speak for R-Ladies Miami

* seeing my "Asking for help online" content re-used in Sharla Gelfand's ["make a reprex... please"](https://twitter.com/sharlagelfand/status/1365665149063987201){target="_blank"} presentation

* developing a unit testing workshop with Gordon Shotwell ???(maybe)

During my curating week I tried to embody the tweets that I value: honest questions, thoughtful discussion, generous sharing, supportive community, and humorous exchanges. And to borrow from Vicki Boykis in the rstudio::global(2021) [keynote](https://rstudio.com/resources/rstudioglobal-2021/your-public-garden/){target="_blank"}, I learned how to create my own public garden. 

# Acknowledgements

