---
title: "Regular, intentional, and time-boxed yak shaving."
description: | 
  A productive struggle towards deep learning.
base_url: https://www.pipinghotdata.com/
twitter:
  site: "@PipingHotData"
  creator: "@PipingHotData"
date: 2022-02-21
author:
  - name: Shannon Pileggi
output:
  distill::distill_article:
    toc: true
    toc_float: true
    toc_depth: 1
    self_contained: false
draft: false
creative_commons: CC BY
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(tidyverse)
library(tweetrmd)
```

# TL; DR

# Background

When Jenny Bryan tweeted this a year ago, I thought about it. For a long time.

```{r}
tweetrmd::include_tweet("https://twitter.com/JennyBryan/status/1366107942487949312")
```

Jenny made the process of yak shaving sound so logical, structured, and easy. 
I had the wonderful opportunity to discuss this with Jenny; here is a summary
of our discussion.

# Yak shaving


From [Jessie Kerr](https://jessitron.com/2017/07/03/a-taxonomy-of-yak-shaving/){target="_blank"}

> Yak Shaving is: doing seemingly-unrelated tasks to get some real task done.


I had never heard of a yak shave before, so I read a lot of people's takes on it.
For some, the yak shave can have negative connotations as a time suck that prevents 
a goal from realization. Others, like Jenny and Jessica, frame it as a growth opportunity, 
if well-positioned.

Individuals also classify the yak shave in various ways (like
Jessica Kerr's [taxonomy of yak shaving](https://jessitron.com/2017/07/03/a-taxonomy-of-yak-shaving/){target="_blank"})
and for various purposes (like Elliot Williams's [hacker vs maker mode](https://hackaday.com/2016/08/09/yak-shaving-hacker-mode-vs-maker-mode/){target="_blank"}). 

To narrow down the scope of the yak shave for this this post, we are discussing
fundamentally reactive processes (not proactive or creative). Consider:

* Learning how complicated tasks mostly out of our control interact with each other. 

* Adopting a more productive debugging mindset than, "Gosh, R is weird." 

* Iterating on workflows that can be improved. 

# Intentional

What items make for a good yak shave? Consider tackling items of high frequency
and those that have painful workarounds. Solving ubiquitous workflow problems that 
permeate all of your projects can have great impact. If you want to be <process oriented>
about it, document notes about hang ups or things you do not understand. This can 
provide evidence of tasks that repeatedly cause friction in workflows or items 
where deeper understanding could alleviate pain points. 

# Regular

So what is the cadence of regular? Not all yaks are equal in size - think about a
mini yak on most days, a medium yak most weeks, and a large 
yak once a month. 

Of course, this depends on the competing demands on your brain resources. Be kind
to  yourself. When you are against a deadline or are mulling over a complex problem, 
you are unlikely to have the mental capacity to successfully and efficiently tackle
the yak shave at hand. 

Instead, consider approaching a yak shave after you have completed a big chunk of
work and you are settling into the next thing. Often it is something you found a
workaround for in a recent project, but now you are circling back for a more 
proper exploration and, if you are lucky, solution.

<aside>
This sounds similar in spirit to [friction logs](https://www.trychameleon.com/blog/friction-logs){target="_blank"} that Alison Hill advocates for good product management, though perhaps not as in depth.
</aside>

# Time-boxed 

Time boxing can be explicit. For example, today, I will give myself 1 hour to research
X before I have to move on to another task. Time boxing can also be more flexible 
depending on the situation and task.

Don't let perfection defeat your time box. If you find a solution that lets
you move forward, even if it is not satisfactory, it is OK to take the
easy way out until you have the mental space to circle back.

# Relection

<not sure if this fits>
One time I heard a talk where someone said, "Deep learning requires a productive
struggle." That really stuck with me. I frequently feel like I am struggling to figure
out new systems or write better code, and it does not always feel productive in the moment.
It is nice to know that others feel the same, and it is normal.

My efforts at yak shaving usually consist of documentation deep dives, 
mad Google searches, and seemingly endless combinations of code that
make me want to throw my computer out the window. Sometimes the stars align
and this works. Sometimes it does not. 

More recently, after I do my initial due diligence with documentation, searching, 
and coding, I weigh the time it takes to post a question online (~30 minutes,
depending on the complexity) versus the time I think it might take me to solve the 
problem alone. If I have the luxury to wait a few days for an answer, posting a 
question online often saves me _hours_ of trial and error with input from those 
more experienced in the domain.

Until I achieve the bliss of incremental learning. 🧘

Every day I strike a balance between incremental yak shaves with reality and
time constraints. Some days I simply maintain the status quo, and some days I move forward 
towards that incremental change. And when I do, it is worth it. 

<Maybe a different closer>
As Jessica Kerr [shared](https://jessitron.com/2021/02/27/one-key-trick-to-becoming-a-wizard-developer/), 
broadening your knowledge through yak shaving can come in handy at unpredictable times. 


