---
title: "Regular, intentional, and time-boxed yak shaving."
description: | 
  Productive incremental learning.
base_url: https://www.pipinghotdata.com/
twitter:
  site: "@PipingHotData"
  creator: "@PipingHotData"
date: 2022-02-21
author:
  - name: Shannon Pileggi
output:
  distill::distill_article:
    toc: true
    toc_float: true
    toc_depth: 1
    self_contained: false
draft: false
creative_commons: CC BY
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(tidyverse)
library(tweetrmd)
```

# TL; DR

# Background

Jenny Bryan, a Software Engineer at RStudio, is known for developing R packages
and educational materials, especially as they relate to reducing friction in workflows. 
A year ago, Jenny tweeted advice regarding the concept of "yak shaving" that made
self-improvement sound so logical, structured, and easy. 

```{r}
tweetrmd::include_tweet("https://twitter.com/JennyBryan/status/1366107942487949312")
```

I thought about it for a long time, and I had questions. I had the wonderful 
opportunity to discuss this with Jenny; here is a summary.

# Yak shaving


I had never heard of a yak shave before, so I read a bit about it.
Yak shaving is a term used frequently in software engineering; a [definition from Jessie Kerr](https://jessitron.com/2017/07/03/a-taxonomy-of-yak-shaving/){target="_blank"} is

> Doing seemingly-unrelated tasks to get some real task done.

Read Jessica's post [One key trick to becoming a wizard developer](https://jessitron.com/2021/02/27/one-key-trick-to-becoming-a-wizard-developer/){target="_blank"}
for an example of a yak shave.

For some, the yak shave can have negative connotations as a time suck that prevents 
a goal from realization. Others, like Jenny and Jessica, frame it as a growth opportunity, 
if well-positioned.

Individuals also classify the yak shave in various ways (like
Jessica Kerr's [taxonomy of yak shaving](https://jessitron.com/2017/07/03/a-taxonomy-of-yak-shaving/){target="_blank"} 
with five types of yaks) and for various purposes (like Elliot Williams's [hacker vs maker mode](https://hackaday.com/2016/08/09/yak-shaving-hacker-mode-vs-maker-mode/){target="_blank"}). 

To narrow down the scope of the yak shave for this this post, we are discussing
a fundamentally reactive processes (not proactive or creative). Consider:

* Learning how complicated tasks mostly out of our control interact with each other. 

* Iterating on workflows that can be improved. 

* Adopting a more productive debugging mindset than, "Gosh, R is weird." 

# Intentional

Good candidates for a yak shave include tasks

* of high frequency, 

* with tedious workarounds, or

* where deeper understanding could alleviate pain points.

If you want to be process oriented about it, document notes about items and workflows
that cause friction, which can provide evidence of priority items. Solving pain points 
that permeate all of your projects can have great impact. 

<aside>
This sounds similar in spirit to [friction logs](https://www.trychameleon.com/blog/friction-logs){target="_blank"} that Alison Hill advocates for good product management, though perhaps not as in depth.
</aside>

# Regular

Yaks differ in size and scope. For the cadence of regular, consider
a mini yak on most days, a medium yak most weeks, and a large yak once a month. 

Of course, this depends on the competing demands of your brain resources. Be kind
to  yourself. When you are against a deadline or are mulling over a complex problem, 
you are unlikely to have the mental capacity to successfully and efficiently tackle
the yak shave at hand. 

Instead, consider approaching a yak shave after you have completed a big chunk of
work and you are settling into the next thing. Often it is something you found a
workaround for in a recent project, but now you are circling back for a more 
proper exploration and, if you are lucky, solution.


# Time-boxed 

Time boxing depends on the situation and task. For mini yaks, you might be able 
to make great progress and find a satisfactory solution within an hour. For larger
yaks, this might be something you return to several times. 

Jenny and I both tackle these kinds of things at night or on the weekend. 
Puttering on the computer outside of prime work hours helps us not be too
fussed about whether experiments pan out, and forces us to think about it a little,
drop it, sleep, do other stuff, and pick up again. Sometimes you get ideas
while you are doing something else or you see something new when you return
with fresh eyes.

Don't let perfection defeat your time box. If you find a solution that lets
you move forward, even if it is not satisfactory, it is OK to take the
"easy" way out until you have the mental space to circle back.

# Relection


My efforts at yak shaving usually consist of documentation deep dives, 
Google searches, and seemingly endless combinations of code that
mostly get tossed out. Sometimes the stars align and this works. 
Sometimes it does not. 

More recently, after I do my initial due diligence with documentation, searching, 
and coding, I weigh the time it takes to post a question online (~30 minutes,
depending on the complexity) versus the time I think it might take me to solve the 
problem alone. If I have the luxury to wait a few days for an answer, posting a 
question online often saves me _hours_ of trial and error with input from those 
more experienced in the domain, helping me shave that yak faster.

Take some (but not all!) opportunities to yak shave productively on a regular basis.
Struggling to figure out new systems or write better code does not always feel 
productive in the moment, but all learning is good learning, even if you are learning
what doesn't work. 

<aside>
Repeat after me: All learning is good learning. All learning is good learning. All learning is good learning.
</aside>


There is a balance to strike between yak shaves and reality / time constraints. 
Some days I simply maintain the status quo, and some days I move forward 
towards that productive incremental learning. And when I do, it is worth it.

-> As Jessica Kerr [shared](https://jessitron.com/2021/02/27/one-key-trick-to-becoming-a-wizard-developer/), 
broadening your knowledge through yak shaving can come in handy at unpredictable times. 

-> Self improvement

-> Personal satisfaction

-> More efficient future you

