---
title: "Regular, intentional, and time-boxed yak shaving."
description: | 
  Incremental deeper learning.
base_url: https://www.pipinghotdata.com/
twitter:
  site: "@PipingHotData"
  creator: "@PipingHotData"
date: 2022-02-21
author:
  - name: Shannon Pileggi
output:
  distill::distill_article:
    toc: true
    toc_float: true
    toc_depth: 1
    self_contained: false
draft: false
creative_commons: CC BY
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(tidyverse)
library(tweetrmd)
```

# TL; DR

# Background

When Jenny Bryan tweeted this a year ago, I thought about it. For a long time.

```{r}
tweetrmd::include_tweet("https://twitter.com/JennyBryan/status/1366107942487949312")
```

Jenny made the process of yak shaving sound so logical, structured, and easy. My 
efforts to become a better programmer usually make me want to throw my computer out the
window with bouts of documentation deep dives and mad Google searches. Until I 
achieve the bliss of incremental learning. 🧘

I had the wonderful opportunity to discuss this with Jenny; here is a summary
of our discussion.

# Yak shaving


From [Jessie Kerr](https://jessitron.com/2017/07/03/a-taxonomy-of-yak-shaving/){target="_blank"}

> Yak Shaving is: doing seemingly-unrelated tasks to get some real task done.


I had never heard of a yak shave before, so I read a lot of people's takes on it.
For some, the yak shave can have negative connotations as a time suck that prevents 
a goal from realization. Others, like Jenny and Jessica, frame it as a growth opportunity, 
if well-positioned.

For this this article, we are discussing a fundamentally reactive (not proactive 
or creative) processes. Learning how complicated tasks mostly out of our control
interact with each other. Adopting a more productive debugging mindset than, 
"Gosh, R is weird." Iterating on workflows that can be improved. 


# Regular

Take opportunities to yak shave productively on a regular basis. Not all yaks are 
equal in size - think about a mini yak on most days, a medium yak most weeks, and
a large yak once a month. Of course, this depends 

When you are against a deadline or have a complicated problem loaded into your brain,
this is not the best time to do this. Consider approaching this after you have 
completed a big chunk of work and you are settling into the next thing. Often it 
is something you found a workaround for in a recent project, but  now now you’re 
circling back for a more proper exploration and, if you’re lucky, solution.

# Intentional

What items make for a good yak shave? Jenny documents "notes to self" with little
hangnails or things you don't understand. That can provide evidence of tasks that
repeatedly cause friction in workflows. Consider tackling items of high frequency
and those that have painful workarounds. Solving workflow problems that affect 
every single project your work on has great impact.


From Jenny: Then eventually I’ll take the time to get to the bottom of it.

<aside>
This sounds similar in spirit to [friction logs](https://www.trychameleon.com/blog/friction-logs){target="_blank"} that Alison Hill advocates for good product management, though perhaps not as in depth.
</aside>

# Time-boxed 

# My practices

strike the balance between incremental elevations / solutions with reality and time constraints. It is ok to not to not have perfect solutions, the key is that you keep moving forward. And some days moving forward is actually maintaining the status quo, and other days moving forward means development and change.

As Jessica Kerr [shared](https://jessitron.com/2021/02/27/one-key-trick-to-becoming-a-wizard-developer/), 
broadening your knowledge through yak-shaving can come in handy at unpredictable times. 
When you can, go forward and dig in a bit.
