---
title: "Code line highlighting"
format: revealjs
---

## Highlight line 1

```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "1"
x <- 1:10
x
LETTERS[x]
```

## Highlight line 2

```{r, echo=TRUE, eval=FALSE, `code-line-numbers`="2"}
x <- 1:10
x
LETTERS[x]
```

## Highlight line 3

```{.r code-line-numbers="3"}
x <- 1:10
x
LETTERS[x]
```




## Explanation

In general, `code-line-numbers` is a Quarto specific code cell option for presentation 
formats of `revealjs`. This means that `code-line-numbers` will not render code
highlighting when used in Quarto documents (`.qmd`) of format `html`, nor within
an R markdown document (`.Rmd`). Note that Quarto chunk options have a dash their name.

For a complete overview of all code cell options supported in Quarto, see
`Quarto -> Reference -> Code Cells -> Knitr`
<https://quarto.org/docs/reference/cells/cells-knitr.html#code-output>

Method 1: 

Here, we are executing code cell options via special comments indicated by `#|`. 
Comment-based syntax is recommended for Quarto to make documents more portable and 
consistent across execution engines. See 
`Quarto -> Guide -> Computations -> Using R -> Chunk Options`

<https://quarto.org/docs/computations/r.html#chunk-options>

Method 2:

Quarto cell options have dash in their name; R only supports dashes in variable 
names when wrapped in the back tick.


Method 3: 

`.r` signals a non-compute code cell (which is what we are using here with `eval=FALSE`). 
These are verbatim code chunks with a class set on them for language highlighting. 
In this case, the language is R.
(Is there a reference for this?)

