---
title: "Ease renv::restore() by updating your repository to Posit Public Package Manager"
description: |
  P3M hosts CRAN binaries of previous releases of R packages
base_url: https://www.pipinghotdata.com/
twitter:
  site: "@PipingHotData"
  creator: "@PipingHotData"
date: 2024-09-13
author:
  - name: Shannon Pileggi
output:
  distill::distill_article:
    toc: true
    toc_float: true
    toc_depth: 2
    self_contained: false
draft: false
creative_commons: CC BY
---

# TL; DR

Updating a `renv` project's package repository from CRAN to P3M allows a user
to more quickly install previous versions of R packages. 

# Background

The `{renv}` package is wonderful for documenting and controlling package versions within 
an R project. See [You should use renv](https://youtu.be/GwVx_pf2uz4?si=agbC-r-us57GF-7f) by E. David Aja at rstudio::conf(2022) for motivation and [Introduction to renv](https://rstudio.github.io/renv/articles/renv.html) for how to get started.

# Package restoration

`{renv}` facilitates restoration of specific versions of packages documented
in the `renv.lock` via `renv::restore()`, which installs packages from
a specified repository. In the `renv.lock`, for many users the repository specification
might look something like this by default:

```
   "Repositories": [
      {
        "Name": "CRAN",
        "URL": "https://cloud.r-project.org"
      }
    ]
```

This will vary user to user based on their specific settings and configurations.
What this means is that `{renv}` is going to that specific url to find and install
packages that are available there. In this case, we are installing from
a cloud instance of CRAN.

However, if the packages documented in the `renv.lock` are of an older version, then the binaries for this package are likely no longer available on CRAN. This means that
when `{renv}` attempts to install that older package from CRAN,
the packages will likely have to compile from source which takes longer and may fail for 
various reasons. This can be a frustrating process to work through.

# Solution

Luckily, the [Posit Public Package Manager (P3M)](https://packagemanager.posit.co/client/#/) exists! This free service hosts binaries of both latest packages and older packages, solving the time-consuming problem of installing from source. Learn more in [Bring Your Own Binary Packages with RSPM (2022)](https://posit.co/blog/publishing-your-own-binary-packages-with-rspm-2022-07/). 

To take advantage, you need to update the repositories
specified in the `renv.lock` - replace the CRAN repository with P3M. You can do manually edit the `renv.lock`, or achieve this via

```{r, eval = FALSE}
renv::lockfile_modify(repos = c(
  P3M = "https://p3m.dev/cran/latest"
  )) |> 
  renv::lockfile_write()
```

which would result in a modified `renv.lock` that looks like this:

```
   "Repositories": [
      {
        "Name": "P3M",
        "URL": "https://packagemanager.posit.co/cran/latest"
      }
    ]
```

Note 1: The P3M url specified here works for both Mac and Windows OS (not Linux).
Use the [`SETUP`](https://packagemanager.posit.co/client/#/repos/cran/setup) tool to determine what url to use for your needs.

Note 2: You can specify more than one repository for package installation. For 
example, you can also specify [R-universe](https://r-universe.dev/search) from [ROpenSci](https://ropensci.org/) or your organization's 
Posit Package Manager if you subscribe to that service. That would look something 
like:

```{r, eval = FALSE}
renv::lockfile_modify(repos = c(
  P3M = "https://p3m.dev/cran/latest",
  ropensci = "https://ropensci.r-universe.dev",
  org_rspm = "http://rspm.xxxx.org/xxxREPO/latest"
  )) |> 
  renv::lockfile_write()
```

```
   "Repositories": [
      {
        "Name": "P3M",
        "URL": "https://packagemanager.posit.co/cran/latest"
      },
      {
        "Name": "ropensci",
        "URL": "https://ropensci.r-universe.dev"
      },
      {
        "Name": "org_rspm",
        "URL": "http://rspm.xxxx.org/xxxREPO/latest"
      }
    ]
```

# More generally

Outside the scope of `{renv}` projects, you can set the default repository for
package installation via the `Global Options` menu in RStudio - see the P3M `SETUP` tool for details. Alternatively, you can specify this via options in your R profile:

```
options("repos" = c(
  "P3M" = "https://p3m.dev/cran/latest",
  "ropensci" = "https://ropensci.r-universe.dev",
  "org_rspm" = "http://rspm.xxxx.org/xxxREPO/latest"
  )
```

If these are your default options, then they will be adopted when you initiate a new
`renv` project.

# Conclusion    

If you have not yet taken advantage of the Posit Public Package Manager, I highly
recommend you do so! It can save a lot of time and frustration when working on 
`renv` projects that utilize older versions of packages, and more generally is
a great package installation resource to default to.

# Acknowledgments

Thanks to [E. David Aja](https://edavidaja.com/) for reviewing this post. ❤️