---
title: "Custom interactive sunbursts with ggplot in R"
description: |
  geom_rect() + geom_coord() + {ggiraph}
base_url: https://www.pipinghotdata.com/
preview: 
twitter:
  site: "@PipingHotData"
  creator: "@PipingHotData"
date: 06-01-2021
author:
  - name: Shannon Pileggi
output:
  distill::distill_article:
    toc: true
    toc_depth: 1
    self_contained: false
draft: false
creative_commons: CC BY
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# TL; DR

I manually defined coordinates for [`geom_rect()`](https://ggplot2.tidyverse.org/reference/geom_tile.html){target="_blank"} to create custom sunbursts with [`ggplot()`](https://ggplot2.tidyverse.org/reference/ggplot.html){target="_blank"}. Pair this with [`{ggiraph}`](https://davidgohel.github.io/ggiraph/index.html){target="_blank"} package to create an awesome interactive graphic!


# Background

Sunbursts in R first caught my eye from the [`sunburstR`](https://cran.r-project.org/web/packages/sunburstR/vignettes/sunburst-2-0-0.html){target="_blank"} package with its amazing interactivity through `d3`. However, I had trouble customizing it and sought a `ggplot` alternative. You can create a version using `geom_col`, as shown in this [RStudio Community Post](https://community.rstudio.com/t/how-to-create-a-ggplot2-pie-and-donut-chart-on-same-plot/42798/3){target="_blank"}; I chose to go the more complex `geom_rect` route for more customization options. I was pretty excited when I got it to work!

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">pretty pumped I got this custom &quot;sunburst&quot;-esque figure to work!! <a href="https://twitter.com/hashtag/ggplot?src=hash&amp;ref_src=twsrc%5Etfw">#ggplot</a> <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a> <a href="https://t.co/L88IQWoVCs">pic.twitter.com/L88IQWoVCs</a></p>&mdash; Shannon Pileggi (@PipingHotData) <a href="https://twitter.com/PipingHotData/status/1215699305555746816?ref_src=twsrc%5Etfw">January 10, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

# The data

To illustrate the concept on a publicly available data set, I am using the May 2021 [`#TidyTuesday` salary data](https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-05-18/readme.md){target="_blank"}. Here, I'll be looking at the top 5 industries and their top 5 job titles by number of respondents. 

The data attributes that work well for this visualization include:

1. 4 to 6 categories, with associated weight or frequency of occurrence

   * our categories are the top 5 industries 

2. 4 to 6 sub-items within each category, with associated weight or frequency of occurrence

   * our sub-items are the top 5 job titles within industry

3. _optional:_ a metric for the sub-items

   * our metric is the median salary for the titles

# Getting started

This material was developed using:

| Software / package  | Version               |
|---------------------|-----------------------|
| R                   | 4.0.5                 | 
| RStudio             | 1.4.1103              | 
| `tidyverse`         | 1.3.1                 |
| `ggiraph`           | 0.7.8                 |
| `glue`              | 1.4.2                 |
| `pals`              | 1.6                   |


```{r}
library(tidyverse) # general use
library(ggiraph)   # interactive ggplots
library(glue)      # pasting strings
library(pals)      # color palette

```

# Data processing

## Import

First, import the data from GitHub.

```{r}
# import data
survey_raw <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-05-18/survey.csv')
```

## Summarize

Next, summarize the data in terms of the top 5 industries and top 5 job titles. For the job title, we'll get (1) the percent of respondents with that title, and (2) the median salary for that title.

```{r}
survey_summary <- survey_raw %>% 
  # identify top industries
  group_by(industry) %>% 
  mutate(
    # number of respondents per industry
    n_industry = n()
  ) %>%
  ungroup() %>% 
  arrange(desc(n_industry), industry) %>%
  mutate(
    # rank of industry in terms of number of survey respondents
    # know a better way to do this bit? pls let me know!
    n_industry_rank = as.numeric(factor(-n_industry)),
    # very light cleaning of job title
    job_title = tolower(job_title) %>% 
      str_remove_all("[:digit:]+") %>% 
      str_trim() 
    ) %>% 
  # reduce to top 5 industries present
  filter(n_industry_rank %in% 1:5) %>% 
  # identify top 5 job titles in each industry 
  group_by(n_industry_rank, n_industry, industry, job_title) %>%
  summarize(
    # number of titles within industry
    n_title = n(),
    # median salary per title
    median_salary = median(annual_salary)
  ) %>% 
  ungroup() %>% 
  arrange(n_industry_rank, industry, desc(n_title)) %>% 
  group_by(industry) %>% 
  # keep top 5 job titles within industry
  slice(1:5) %>%
  # recompute number of respondents per industry based on titles selected
  mutate(n_industry = sum(n_title)) %>% 
  ungroup() %>% 
  mutate(
    # relative scale of salary for plotting, on 0 - 100 scale
    median_salary_scaled = median_salary / max(median_salary) * 100,
    # base n respondents
    n_total = sum(n_title),
    # percent of respondents per title
    pct_title = n_title / n_total * 100,
    pct_industry = n_industry / n_total * 100,
    # unique job title, as some repeat across industry
    title_unique = glue::glue("{str_sub(industry, 1, 3)}: {job_title}")
  ) %>% 
  arrange(desc(pct_industry, pct_title))
```

Here is a preview of our data, which is the basis needed for the sunburst's custom rectangles.

```{r}
# view input data
survey_summary %>% 
  dplyr::select(industry, n_industry, pct_industry, job_title, n_title, pct_title, median_salary, median_salary_scaled) %>% 
  mutate_at(vars(pct_industry,pct_title, median_salary_scaled), scales::number, accuracy = 0.1) %>% 
  DT::datatable()
```

## Assign colors

Pick a color palette for industry, and name the color vector.

```{r}
colors_industry <- pals::cubicyf(n = 5) %>% 
  set_names(unique(survey_summary[["industry"]]))
```

```{r}
colors_industry
```

## Outer ring

Next, I manually define rectangles for the outer ring of the sunburst. I figured out the bits to create custom rectangles from various blog posts that I did not save. It could have been this post on [variable width column charts](https://learnr.wordpress.com/2009/03/29/ggplot2-variable-width-column-chart/){target="_blank"}. I got tips on how to angle text for labels from the [R Graph Gallery's circular barplots](http://www.r-graph-gallery.com/296-add-labels-to-circular-barplot.html){target="_blank"}; sample code is shown below but not implemented in the figure. 

The idea of the rectangles' color assignment is that darker shades represent more frequent or heavier weight items within a category. To create the colors, I concocted a solution to implement a degree of transparency off of the original industry colors based on the total number of sub-categories. Jake Riley shared with me [an alternative solution for sub-category color assignment](https://stackoverflow.com/questions/56120815/how-to-factor-sub-group-by-category){target="_blank"} that he contributed on Stack Overflow.

```{r}
ring_outer <- survey_summary %>% 
  # join colors to data frame ----
    left_join(enframe(colors_industry, name = "industry", value = "color_industry"), 
            by = "industry") %>%
  # arrange descending industry but ascending title
  # to compute title colors
  arrange(desc(n_industry), industry, n_title, title_unique) %>% 
  group_by(industry) %>% 
  mutate(
    # enumerate title in industry
    id_title = row_number(),
    # number of titles per industry
    # this can vary - it does not have to be same number per category
    num_title = max(id_title)
  ) %>% 
  ungroup() %>% 
  mutate(
    # degree of transparency based on number of attributes ----
    color_trans = id_title / num_title ,
    color_title = map2_chr(color_industry, color_trans, ~ adjustcolor(.x, .y))
  ) %>% 
  # this arrange specifies the ordering of the figure
  # clockwise ---
  arrange(-pct_industry, industry, -pct_title, title_unique) %>%
  # counter clockwise ---
  # arrange(pct_industry, industry, pct_title, title_unique) %>% 
  mutate(
    industry = fct_inorder(industry),
    title_unique = fct_inorder(title_unique)
  ) %>% 
  group_by(industry) %>% 
  mutate(cum_title_pct = cumsum(pct_title)) %>% 
  ungroup() %>% 
  mutate(
    cum_industry_pct = cumsum(cum_title_pct),
    cum_all = cumsum(pct_title)
  ) %>% 
  mutate(
    # compute coordinates of the rectangles ----
    # the 0.3 is an adjustment on the 0 to 100 scale to add a small
    # amount of white space between rectangles
    rect_x_max = cumsum(pct_title) - 0.3,      # xmax ----
    rect_x_min = rect_x_max - pct_title + 0.3, # xmin ----
    rect_x_mid = (rect_x_min + rect_x_max)/2,  # xmid ----
    # angles in case adding text to plot, not shown in figures below ----
    angle = 90 - 360 * rect_x_mid / 100,
    hjust = ifelse(angle < -90, 1, 0),
    angle = ifelse(angle < -90, angle + 180, angle),
    # label for ggiraph
    label_title = glue::glue('{scales::percent(pct_industry/100, accuracy = 1)} ({n_industry}/{n_total}) {industry}\n {scales::percent(pct_title/100, accuracy = 1)} ({n_title}/{n_total}) {job_title} \n {scales::dollar(median_salary)} median salary'))
```

```{r}
# view sunburst data
ring_outer %>% 
   mutate_at(vars(pct_industry,pct_title, median_salary_scaled, rect_x_max, rect_x_min), scales::number, accuracy = 0.1) %>% 
  dplyr::select(industry, n_industry, pct_industry, job_title, n_title, pct_title, id_title, num_title, color_trans, color_title, rect_x_max, rect_x_min, median_salary, median_salary_scaled) %>% 
DT::datatable()
```

## Combine colors

Combine all industry and title colors in a single vector.

```{r}
colors_title <- ring_outer[["color_title"]] %>% 
  set_names(ring_outer[["title_unique"]])

all_colors <- c(colors_industry, colors_title)
```

## Inner ring

Lastly in data prep, we create a smaller data set of rectangle coordinates just for the industries. The requires a dummy extra row to create white space between the first and last categories.


```{r}
# create an inner ring for industry ----
# there is probably a better way to do this too, please let me know 
# if you have suggestions to streamline! 
ring_inner <- ring_outer %>% 
  dplyr::select(industry, pct_industry, n_industry, n_total, rect_x_max, rect_x_min) %>% 
  # first, get stopping point for each rectangle from least common item ----
  arrange(industry, desc(rect_x_max)) %>% 
  group_by(industry) %>%
  # keep one entry per industry ----
  slice(1) %>% 
  ungroup() %>% 
  dplyr::select(industry, pct_industry, n_industry, n_total, stop = rect_x_max) %>% 
  # second, get starting point for each rectangle from most common item ----
  left_join(
    ring_outer %>% 
      dplyr::select(industry, pct_industry, n_industry, n_total, rect_x_max, rect_x_min) %>% 
      arrange(industry, rect_x_max) %>% 
      group_by(industry) %>% 
      slice(1) %>% 
      ungroup() %>% 
      dplyr::select(industry, pct_industry, n_industry, n_total, start = rect_x_min),
    by = c("industry", "pct_industry", "n_industry", "n_total")
  ) %>% 
  # insert blank row for extra white space where circle starts/stops ---
  bind_rows(
    tibble(
      attr_category = NA, 
      start = 99.7, 
      stop = 100.1)
  ) %>% 
  mutate(
    # compute midpoint for labeling if needed ----
    mid = (stop - start)/2 + start,
    # a label for industry pct ----
    label_industry = ifelse(is.na(industry), NA, glue::glue('{scales::percent(pct_industry/100, accuracy = 1)} ({n_industry}/{n_total}) {industry} '))
  )
```

```{r}
# view industry rectangles
ring_inner %>%
  dplyr::select(industry, pct_industry, start, stop) %>% 
   mutate_at(vars(pct_industry, start, stop), scales::number, accuracy = 0.1) %>% 
DT::datatable()
```

# Non-interactive sunbursts

## Variable length sunburst

Finally, we can plot! This first figure has variable length bars, which are represented by the scaled median salary. The outer ring goes from a height of 0 up to a height of the scaled median salary, which is 100 here. Adding extra negative space to the `ylim` creates the donut in the middle.

```{r}
# outer ring for job title
ring_outer %>%
  ggplot() +
  geom_rect(
    aes(
      ymin = 1,
      # variable length bars
      ymax = median_salary_scaled,
      xmin = rect_x_min,
      xmax = rect_x_max,
      fill = title_unique
    ),
    show.legend = FALSE
  ) +
  coord_polar() +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none",
  ) +
  # fiddle lower limit here to control donut size
  ylim(-40, 100) +
  scale_fill_manual(values = all_colors) +
  # inner industry ring ----
  geom_rect(data = ring_inner,
          aes(
            # the inner ring spans a height from 0 to -20
            ymin = -20,
            ymax = 0,
            xmin = start,
            xmax = stop,
            fill = industry
          ))
```

Well, it is an example! Not quite as aesthetically appealing as my original example shown in the tweet, but it will do to put code to the page. 

Although I think this figure can be eye-catching, it is challenging to make it informative with labeling and still maintain appeal. For this reason, I have omitted labels.

## Fixed length sunburst

For a simpler alternative, drop the job title metric of median salary and only show the frequency of industry and title in survey respondents by fixing the rectangle height. This figure can effectively communicate parts of a whole.

```{r}
# outer ring for job title
ring_outer %>%
  ggplot() +
  geom_rect(
    aes(
      ymin = 1,
      # fix rectangle height here
      ymax = 20, 
      xmin = rect_x_min,
      xmax = rect_x_max,
      fill = title_unique
    ),
    show.legend = FALSE
  ) +
  coord_polar() +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none",
  ) +
  # fiddle lower limit here to control donut size
  # reduce max lim as well
  ylim(-40, 30) +
  scale_fill_manual(values = all_colors) +
  # inner industry ring ----
  geom_rect(data = ring_inner,
          aes(
            # the inner ring spans a height from 0 to -20
            ymin = -20,
            ymax = 0,
            xmin = start,
            xmax = stop,
            fill = industry
          ))
# could add reference line of some sort
# geom_hline(yintercept = 50, color = "gray", lty = 2)
```


# Interactive sunburst

With some very small tweaks to incorporate the {ggiraph} package, we can transform this to an interactive sunburst!

## Variable length sunburst


```{r}
# outer ring for job title
gg_interactive_variable <- ring_outer %>%
  ggplot() +
  # change 1: use geom_rect_interactive function from ggiraph
  ggiraph::geom_rect_interactive(
    aes(
      ymin = 1,
      # variable length bars
      ymax = median_salary_scaled,
      xmin = rect_x_min,
      xmax = rect_x_max,
      fill = title_unique,
      # change 2: add id for hover over tooltip ---
      tooltip = label_title
    ),
    show.legend = FALSE
  ) +
  coord_polar() +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none",
  ) +
  # fiddle lower limit here to control donut size
  # reduce max lim as well
  ylim(-40, 100) +
  scale_fill_manual(values = all_colors) +
  # change 3: add interactivity to inner ring  ----
  ggiraph::geom_rect_interactive(
    data = ring_inner,
          aes(
            # the inner ring spans a height from 0 to -20
            ymin = -20,
            ymax = 0,
            xmin = start,
            xmax = stop,
            fill = industry,
            # change 4: add tooltip to inner ring
            tooltip = label_industry
          ))
```

```{r}
# change 5: create interactive version using ggirafe
girafe(ggobj = gg_interactive_variable,
       # this option fills tooltip background with same color from figure
       options = list(opts_tooltip(use_fill = TRUE)))
```


## Fixed length sunburst


```{r}
# outer ring for job title
gg_interactive_fixed <- ring_outer %>%
  ggplot() +
  # change 1: use geom_rect_interactive function from ggiraph
  ggiraph::geom_rect_interactive(
    aes(
      ymin = 1,
      # fix rectangle height here
      ymax = 20, 
      xmin = rect_x_min,
      xmax = rect_x_max,
      fill = title_unique,
      # change 2: add id for hover over tooltip ---
      tooltip = label_title
    ),
    show.legend = FALSE
  ) +
  coord_polar() +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none",
  ) +
  # fiddle lower limit here to control donut size
  ylim(-40, 30) +
  scale_fill_manual(values = all_colors) +
  # change 3: add interactivity to inner ring  ----
  ggiraph::geom_rect_interactive(
    data = ring_inner,
          aes(
            # the inner ring spans a height from 0 to -20
            ymin = -20,
            ymax = 0,
            xmin = start,
            xmax = stop,
            fill = industry,
            # change 4: add tooltip to inner ring
            tooltip = label_industry
          ))
```

```{r}
# change 5: create interactive version using ggirafe
girafe(ggobj = gg_interactive_fixed,
       # this option fills tooltip background with same color from figure
       options = list(opts_tooltip(use_fill = TRUE)))
```

# Fin

Like this figure? Use it in your own work? Or have suggestions for improvement in either code or the graphic? Please, let me know!

# For fun

I did experiment with some variations on this figure that did not make the cut!

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">Monday&#39;s work 🤔🥰 <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a> <a href="https://twitter.com/hashtag/ggplot?src=hash&amp;ref_src=twsrc%5Etfw">#ggplot</a> <a href="https://t.co/GbZWLhLeDA">pic.twitter.com/GbZWLhLeDA</a></p>&mdash; Shannon Pileggi (@PipingHotData) <a href="https://twitter.com/PipingHotData/status/1224700519496765440?ref_src=twsrc%5Etfw">February 4, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>


